MS Technical Paper: \\ Placement Algorithms for Heterogeneous FPGAs

### 2025-05-07 latex/paper/sections/08-netlist.tex
netlist stuff

\section{What is a Netlist?}
\label{sec:netlist}
In its most general form, a netlist is a list of every component in an electronic design paired with a list of nets they connect to. 
Depending on the working abstraction level, these components can be transistors, logic gates, macrocells, or increasingly higher-level modules. 
Generally, a net denotes any group of two or more interconnected components.
In an electronics context, a net can be though of as a wire connecting multiple pins between multiple components, with each wire having one voltage source and one or more voltage sinks. 
Thus, one could express the netlist as a hypergraph, nodes representing component ports, hyperedges representing wires connecting two or more component ports. 
More precisely, these hyperedges connect the pins between the components, not the components themselves, with each component exposing multiple pins or ports.

In FPGA context, the components are logical cells (\texttt{LUTs}, \texttt{CARRY4s}, etc.) or hierarchical cells (Verilog module instances) with pins connected together by wires. 
In Vivado, a Netlist can be synthesized as a Hierarchical or a Flattened netlist. 
Figure ~\ref{fig:hierarchical_design} shows an example a Verilog design using module instance hierarchy, while figure ~\ref{fig:hier_netlist} shows the design synthesized into a hierarchical netlist with \textbf{hierarchical cells} and \textbf{leaf cells}. 
The synthesizer attempts to construct the module hierarchy as close to the module instantiation hierarchy defined by the user design entry. 
Figure ~\ref{fig:flat_netlist} shows the same design but synthesized into a flattened netlist.

@@

\end{multicols}
{
    \raggedright
    \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/top_level.png}
    \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_0.png}
    \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_1.png}
    \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_2.png}
    \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_3.png}
    \captionof{figure}{A simple HDL design with module hierarchy.}
    \label{fig:hierarchical_design}
}
{
    \centering
    \includegraphics[valign=c, width=11.5cm]{figures/netlist_synth/hier_netlist.png}
    \includegraphics[valign=c, width=6cm]{figures/netlist_synth/hier_graph.png}
    \captionof{figure}{
        \textbf{Left:} A hierarchical netlist consisting of LUTs and FFs.
        \textbf{Right:} The cell hierarchy graph.
    }
    \label{fig:hier_netlist}
}
\vspace{0.5cm}
{
    \centering
    \includegraphics[valign=c, width=13cm]{figures/netlist_synth/flat_netlist.png}
    \includegraphics[valign=c, width=4cm]{figures/netlist_synth/flat_graph.png}
    \captionof{figure}{
        \textbf{Left:} A flattened netlist consisting of LUTs and FFs.
        \textbf{Right:} The flattened cell hierarchy graph.
    }
    \label{fig:flat_netlist}
}
\begin{multicols}{2}

@@

\begin{itemize}
\item

\end{itemize}

@@

// Traverse the set of nets
for (EDIFNet net : netSet) {
    System.out.println("Net: " + net.getName());
    // Access the ports connected to this net
    Collection<EDIFPortInst> epis = net.getPortInsts();
    for (EDIFPortInst epi : epis) {
        // Access the cell that this port belongs to
        EDIFCellInst eci = epi.getCellInst();
        if (eci == null) {
            // (top_level ports have no associated cell)
            continue;
        } else {
            System.out.println(
                "\tCell: " + eci.getName() + 
                ",\tCellType: " + eci.getCellName()
            );
        }
    }
}

\end{lstlisting}

{
    \centering
    \includegraphics[valign=c, width=\columnwidth]{figures/traversal.png}
    \captionof{figure}{\textttt{Netlist} traversal via \texttt{EDIFHierCellInst}, \texttt{EDIFHierPortInst}, and \texttt{EDIFHierNet} objects}
    \label{fig:traversal}
}

>>>>

\subsection{What is a Netlist?}
\label{subsec:netlist}
In its most general form, a netlist is a list of every component in an electronic design paired with a list of nets they connect to. 
Depending on the abstraction level at hand, these components can be transistors, logic gates, macrocells, or increasingly higher-level modules. 
Generally, a net denotes any group of two or more interconnected components.
In an electronics context, a net can be though of as a wire connecting multiple pins between multiple components, with each wire having one voltage source and one or more voltage sinks. 
Thus, one could express the netlist as a hypergraph, nodes representing components, hyperedges representing wires connecting two or more component. 
More precisely, these hyperedges connect the ports between the components, not the components themselves, with each component exposing multiple ports.

In FPGA context, the components are logical cells (\texttt{LUTs}, \texttt{CARRY4s}, etc.) or hierarchical cells (Verilog module instances) with pins connected together by wires. 
In Vivado, a Netlist can be synthesized as a Hierarchical or a Flattened netlist. 
Figure ~\ref{fig:hierarchical_design} shows an example a Verilog design with modules instantiated in a hierarchy. 
Figure ~\ref{fig:hier_netlist} shows the design synthesized into a hierarchical netlist with \textbf{hierarchical cells} and \textbf{leaf cells}. 
The synthesizer attempts to construct the module hierarchy as close to the module instantiation hierarchy defined by the user design entry. 
Figure ~\ref{fig:flat_netlist} shows the same design but synthesized into a flattened netlist.

@@

RapidWright represents the logical netlist objects via the \texttt{edif} classes:
\begin{itemize}
    \item \texttt{EDIFNetlist}: The full logical netlist of a \texttt{Design}. 
    \item \texttt{EDIFNet}: Represents a logical net within an \texttt{EDIFNetlist}.
    \item \texttt{EDIFHierNet}: Combines an \texttt{EDIFNet} with a full hierarchical instance name to uniquely identify a net in a netlist.
    \item \texttt{EDIFCell}: Represents a logical cell in an \texttt{EDIFNetlist}. 
    \item \texttt{EDIFCellInst}: Represents an instance of an \texttt{EDIFCell}. 
    \item \texttt{EDIFHierCellInst}: An \texttt{EDIFCellInst} with its hierarchy, described by all the \texttt{EDIFCellInst}s that sit above it within the netlist.
    \item \texttt{EDIFPort}: Represents a port on an \texttt{EDIFCell}. 
    \item \texttt{EDIFPortInst}: Represents an instance of a port on an \texttt{EDIFCellInst}. 
    \item \texttt{EDIFHierPortInst}: Combines an \texttt{EDIFHierPortInst} with a full hierarchical instance name to uniquely identify a port instance in a netlist. 
\end{itemize}

\newpage

\end{multicols}
{
    \raggedright
    \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/top_level.png}
    \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_0.png}
    \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_1.png}
    \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_2.png}
    \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_3.png}
    \captionof{figure}{A simple HDL design with module hierarchy.}
    \label{fig:hierarchical_design}
}
\vspace{0.5cm}
{
    \centering
    \includegraphics[valign=c, width=11.5cm]{figures/netlist_synth/hier_netlist.png}
    \includegraphics[valign=c, width=6cm]{figures/netlist_synth/hier_graph.png}
    \captionof{figure}{
        \textbf{Left:} A hierarchical netlist consisting of LUTs and FFs.
        \textbf{Right:} The cell hierarchy graph.
    }
    \label{fig:hier_netlist}
}
\vspace{0.5cm}
{
    \centering
    \includegraphics[valign=c, width=13cm]{figures/netlist_synth/flat_netlist.png}
    \includegraphics[valign=c, width=4cm]{figures/netlist_synth/flat_graph.png}
    \captionof{figure}{
        \textbf{Left:} A flattened netlist consisting of LUTs and FFs.
        \textbf{Right:} The flattened cell hierarchy graph.
    }
    \label{fig:flat_netlist}
}
\newpage
\begin{multicols}{2}

@@

{
    \centering
    \includegraphics[valign=c, width=\columnwidth]{figures/traversal.png}
    \captionof{figure}{\texttt{Netlist} traversal via \texttt{EDIFHierCellInst}, \texttt{EDIFHierPortInst}, and \texttt{EDIFHierNet} objects}
    \label{fig:traversal}
}
\vspace{0.5cm}

@@

// Traverse the set of nets
for (EDIFNet net : netSet) {
    System.out.println("Net: " + net.getName());
    // Access the ports connected to this net
    Collection<EDIFPortInst> epis = net.getPortInsts();
    for (EDIFPortInst epi : epis) {
        // Access the cell that this port belongs to
        EDIFCellInst eci = epi.getCellInst();
        if (eci == null) {
            // (top_level ports have no associated cell)
            continue;
        } else {
            System.out.println(
                "\tCell: " + eci.getName() + 
                ",\tCellType: " + eci.getCellName()
            );
        }
    }
}
\end{lstlisting}

\newcolumn
